# Project 1

Authors: Meerah Waters and Jack Wetzel

## Introduction

This program is designed to create functions to retrieve and analyze data from the US census. The first portion will involve creating a function similar to the get_acs() function from the tidycensus package. It will take inputs from the user to specify the year, variables, and geographic location for which they wish to get data from. The function will then produce a tibble with data adhering to the specifies parameters. Afterwards, there will be additional functions that will allow the user to analyze the previously created tibble. The first is a summary function which will provide counts of categorical variables and basic statistical measures of the numerical variables. The second is a graphing function which takes the summary funciton as an input and produces a box plot based on specified variables.

## Task 1: Data Processing

1.  First we need to install the necessary R packages that are used in the generic functions in this exercise.

```{r, message=FALSE}
library(tidycensus)
library(jsonlite)
library(httr)
library(tibble)
library(dplyr)
library(tidyr)
library(lubridate)
library(tidyselect)
library(ggplot2)
```

2.  Then we need to call the API (this is the default call)

```{r}
acsData <- "https://api.census.gov/data/2022/acs/acs1/pums?get=PWGTP,AGEP,SEX&for=region:04"
```

3.  Then we need to parse the JSON data and put it in tibble format with the helper function

```{r}
apiHelper <- function(apiCall) {
  acsData <- GET(apiCall)
  parsed <- fromJSON(rawToChar(acsData$content))
  acsDataTib <- as_tibble(parsed)
  names(acsDataTib) <- acsDataTib[1,]
  acsDataTib <- acsDataTib[-1,]
}
```

4.  Now were going to write a function to query the API that will allow the user to specify certain items. The function allows for the individual to select a year, Numerical Variables, Categorical Variables, and the Geographic area through the inputs of the function. How the function is written, causes only one singular "Geography" selection, while the Numerical and Categorical Variables are able to have multiple responses, as these are only the column headers being selected.

```{r}
acsFunction <- function(year = 2022, 
                        numericalVar = c("PWGTP", "AGEP"), 
                        categoricalVar = c("SEX"), 
                        geography = "state",
                        geo_code = "04") {

  ## Checking if function input is valid
  if (year > 2022 | year < 2010) {
    stop("Year must be between 2010 and 2022")
  }
  
  if (is.null(categoricalVar)) {
    stop("One categorical variable must be returned")
  }
  
  if (is.null(numericalVar) | all(numericalVar == "PWGTP")) {
    stop("One numeric variable other than PWGTP must be returned")
  }
  
  if (!(geography %in% c("state", "division", "region"))) {
    stop("Invalid Geography value, must be: (state, division, region)")
  }
  
  if (is.null(geo_code)) {
    geo_code <- "*"
  }
  
  ## Building the URL
  variables <- paste(c(numericalVar, categoricalVar), collapse = ",")
  
  ## Ensures PWGTP is always printed
  if (!("PWGTP") %in% numericalVar) {
    variables <- paste("PWGTP", variables, sep = ",")
  }
  
  url <- paste0("https://api.census.gov/data/",
                year,
                "/acs/acs1/pums?get=",
                variables,
                "&for=",
                geography,
                ":",
                geo_code
                )
  
  acsTibble <- apiHelper(apiCall = url)
  
  ## Making variables numeric
  numericalVar <- numericalVar[!(numericalVar %in% c("JWAP", "JWDP"))]
  acsTibble[numericalVar] <- lapply(acsTibble[numericalVar], as.numeric)
  acsTibble$PWGTP <- as.numeric(acsTibble$PWGTP)
  
  ## Making variables factors
  for (i in 1:length(categoricalVar)) {
    cat_url <- paste0("https://api.census.gov/data/2022/acs/acs1/pums/variables/",
                      categoricalVar[i],
                      ".json")
    cat_var <- categoricalVar[i]
    temp <- GET(cat_url)
    temp_list <- temp$content |> rawToChar() |> fromJSON()
    temp_cat <- temp_list$values$item
    cat_values <- temp_cat[sort(names(temp_cat))]
    
    vec <- list()
    vec <- sapply(acsTibble[cat_var], function(x) cat_values[x])
    vec <- sapply(vec, function(x) if (is.null(x)) NA else x)
    vec_num <- unname(unlist(vec))
    acsTibble[cat_var] <- vec_num
  }
  
  acsTibble <- acsTibble |>
    mutate(across(where(is.character), as.factor))
  
  ## Formatting the time variables as minutes since midnight
  if ("JWAP" %in% colnames(acsTibble)) {
    temp <- GET("https://api.census.gov/data/2022/acs/acs1/pums/variables/JWAP.json")
    temp_list <- temp$content |> rawToChar() |> fromJSON()
    temp_JWAP <- temp_list$values$item
    JWAP_values <- temp_JWAP[sort(names(temp_JWAP))]
    JWAP_values[1] <- 0  ## Sets the first value to 0 as the loop cannot handle the first value
    
    for (i in 2:length(JWAP_values)) {
      time_strings <- strsplit(JWAP_values[[i]], " to ")
      time_1 <- time_strings[[1]][1]
      time_2 <- time_strings[[1]][2]
      
      if (grepl("a\\.m\\.", time_1)) {
        time_form_1 <- gsub("a\\.m\\.", "AM", time_1)
      } else {
        time_form_1 <- gsub("p\\.m\\.", "PM", time_1)
      }
      
      if (grepl("a\\.m\\.", time_2)) {
        time_form_2 <- gsub("a\\.m\\.", "AM", time_2)
      } else {
        time_form_2 <- gsub("p\\.m\\.", "PM", time_2)
      }
      
      time_24_1 <- format(strptime(time_form_1, "%I:%M %p"), format="%H:%M")
      time_24_2 <- format(strptime(time_form_2, "%I:%M %p"), format="%H:%M")
      
      time_parts_1 <- strsplit(time_24_1, ":")[[1]]
      time_parts_2 <- strsplit(time_24_2, ":")[[1]]
      
      hours_1 <- as.numeric(time_parts_1[1])
      hours_2 <- as.numeric(time_parts_2[1])
      minutes_1 <- as.numeric(time_parts_1[2])
      minutes_2 <- as.numeric(time_parts_2[2])
      
      total_time_1 <- hours_1 * 60 + minutes_1
      total_time_2 <- hours_2 * 60 + minutes_2
      
      JWAP_values[i] <- (total_time_1 + total_time_2) / 2
    }
    vec <- list()
    vec <- sapply(acsTibble$JWAP, function(x) JWAP_values[[x]])
    vec <- sapply(vec, function(x) if (is.null(x)) NA else x)
    vec_num <- unname(unlist(vec))
    acsTibble$JWAP <- vec_num
  }
  
  if ("JWDP" %in% colnames(acsTibble)) {
    temp <- GET("https://api.census.gov/data/2022/acs/acs1/pums/variables/JWDP.json")
    temp_list <- temp$content |> rawToChar() |> fromJSON()
    temp_JWDP <- temp_list$values$item
    JWDP_values <- temp_JWDP[sort(names(temp_JWDP))]
    JWDP_values[1] <- 0  ## Sets the first value to 0 as the loop cannot handle the first value
    
    for (i in 2:length(JWDP_values)) {
      time_strings <- strsplit(JWDP_values[[i]], " to ")
      time_1 <- time_strings[[1]][1]
      time_2 <- time_strings[[1]][2]
      
      if (grepl("a\\.m\\.", time_1)) {
        time_form_1 <- gsub("a\\.m\\.", "AM", time_1)
      } else {
        time_form_1 <- gsub("p\\.m\\.", "PM", time_1)
      }
      
      if (grepl("a\\.m\\.", time_2)) {
        time_form_2 <- gsub("a\\.m\\.", "AM", time_2)
      } else {
        time_form_2 <- gsub("p\\.m\\.", "PM", time_2)
      }
      
      time_24_1 <- format(strptime(time_form_1, "%I:%M %p"), format="%H:%M")
      time_24_2 <- format(strptime(time_form_2, "%I:%M %p"), format="%H:%M")
      
      time_parts_1 <- strsplit(time_24_1, ":")[[1]]
      time_parts_2 <- strsplit(time_24_2, ":")[[1]]
      
      hours_1 <- as.numeric(time_parts_1[1])
      hours_2 <- as.numeric(time_parts_2[1])
      minutes_1 <- as.numeric(time_parts_1[2])
      minutes_2 <- as.numeric(time_parts_2[2])
      
      total_time_1 <- hours_1 * 60 + minutes_1
      total_time_2 <- hours_2 * 60 + minutes_2
      
      JWDP_values[i] <- (total_time_1 + total_time_2) / 2
    }
    vec <- list()
    vec <- sapply(acsTibble$JWDP, function(x) JWDP_values[[x]])
    vec <- sapply(vec, function(x) if (is.null(x)) NA else x)
    vec_num <- unname(unlist(vec))
    acsTibble$JWDP <- vec_num
  }
  
  ## Removing geography columns from final tibble
  if ("state" %in% colnames(acsTibble)) {
    acsTibble <- acsTibble |>
      select(-state)
  }
  
  if ("division" %in% colnames(acsTibble)) {
    acsTibble <- acsTibble |>
      select(-division)
  }
  
  if ("region" %in% colnames(acsTibble)) {
    acsTibble <- acsTibble |>
      select(-region)
  }
  
  return(acsTibble)
}
```

Now we will test that the acsFunction works as intended. This will be done by giving it different inputs to ensure that all exceptions are handled properly. More tests were done than shown below throughout the coding process, but the ones below show a good picture.

```{r, error=TRUE}
## Test includes values different than function defaults for each input and tests time variable formatting
acsFunction(year = 2019, categoricalVar = c("HISPEED", "SCH"), numericalVar = c("JWAP"), geography = "region", geo_code = "01")

## Tests function when only some inputs are specified
acsFunction(numericalVar = c("JWMNP", "GASP"), geography = "division", geo_code = "5")

## Tests the checks for valid function inputs
acsFunction(year = 2008)
acsFunction(categoricalVar = NULL)
acsFunction(numericalVar = c("PWGTP"))
acsFunction(geography = "city")
```

```{r}
## Runs the acsFunction with default inputs
defaultTibble <- acsFunction()
```

5.  Now we will build a new function that has similar parameters to the one in step 4 with the exception of allowing the user to specify multiple years. This will call the previous function multiple times and combine multiple tibbles into one. Additionally, it creates a new column titled year.

```{r}
acsMultiYear <- function(years,
                         numericalVar = c("PWGTP", "AGEP"), 
                         categoricalVar = c("SEX"), 
                         geography = "state",
                         geo_code = "04") {
  
  tibbleList <- list()  ## initializing empty list for the loop
  
  for (i in 1:length(years)) {
    singleYrTib <- acsFunction(years[i], numericalVar, categoricalVar, geography, geo_code)
    singleYrTib$year <- years[i]  ## creates a new variable based on the year
    tibbleList[[i]] <- singleYrTib
  }
  
  multYrTib <- bind_rows(tibbleList)  ## combines the list of tibbles into a single tibble
  
  return(multYrTib)
}
```

We will now test that the acsMultiYear function works by giving it an input with two different years. Due to high processing time of including multiple years, the head and tail functions were used. 

```{r}
multYrTest <- acsMultiYear(years = c(2018, 2021))
head(multYrTest)  ## year column is 2018
tail(multYrTest)  ## year column is 2021
```

## Task 2: Data Summary

6.  In this step, we are creating a summary function of our overall tibble, while also adding a new class to the overall data. The summary is dependent on the numerical and categorical variables selected in the acsFunction previously. By default, the function will summarize all variables present in the given tibble, but users can specify the only summarize specific ones.

```{r}
acsSummary <- function(classcensus = defaultTibble,
                       numericalVar = NULL,
                       categoricalVar = NULL) {
    
    ## Adding the census class to the tibble
    class(classcensus) <- c("census", class(classcensus))
    
    ## Checks if user specifies variables
    if (!is.null(numericalVar)) {
      classcensus <- classcensus |>
        select(where(is.factor), PWGTP, all_of(numericalVar))
    }
    
    if (!is.null(categoricalVar)) {
      classcensus <- classcensus |>
        select(where(is.numeric), all_of(categoricalVar))
    }
    
    ## Creates a tibble with counts for categorical variables
    acsSummaryCat <- classcensus |>
      pivot_longer(where(is.factor), names_to = "column_name", values_to = "var_count") |>
      count(PWGTP, column_name, var_count) |>
      pivot_wider(names_from = c(column_name, var_count), values_from = n, values_fill = 0)
    
    ## Creates a tibble with weighted mean & sd of numerical variables
    acsSummaryNum <- classcensus |>
      group_by(PWGTP) |>
      mutate(across(where(is.numeric), 
                    list(mean = mean, sd = sd), 
                    .names = "{.col}_{.fn}")) |>
      arrange(PWGTP) |>
      select(PWGTP, ends_with("mean"), 
             ends_with("sd")) |>
      distinct()
    
    ## Combines the numerical and categorical tibbles
    acsTibbleSummary <- full_join(acsSummaryCat, acsSummaryNum)
      
    return(acsTibbleSummary)
}
```

Now we will test that the acsSummary function works as intended. This test makes sure that there are columns for the categorical variable counts along with a column for the mean & sd of numerical variables. It also tests that users are able to specify to summarize only specific variables from the original tibble.

```{r}
test_tibble <- acsFunction(2021, categoricalVar = c("HHL", "FER"), numericalVar = c("JWAP", "GASP"))
testsummary <- acsSummary(classcensus = test_tibble, numericalVar = c("JWAP"), categoricalVar = ("HHL"))
testsummary
```

7.  After creating our summary of our selected data, we are now able to plot the data in a digestible graph!

```{r}
acsPlot <- function(dataset = testsummary,
                    numericalVar = "mean_GASP",
                    categoricalVar = "mean_HHL") {
 
  ggplot(dataset,
         aes(x = get(categoricalVar), y = get(numericalVar), weight = PWGTP)) +
    geom_boxplot()
}
```

```{r}
acsPlot(testsummary, "JWAP_sd", "HHL_Spanish")
```