Project 1

Objective: Create generic functions to automatically summarize and plot certain returned data.

Authors: Meerah Waters and Jack Wetzel

## Task 1: Data Processing

1.  First we need to install the necessary R packages that are used in the generic functions in this exercise.

    ```{r}
    library(tidycensus)
    library(jsonlite)
    library(httr)
    library(tibble)
    library(dplyr)
    library(tidyr)
    library(lubridate)
    library(tidyselect)
    library(ggplot2)
    ```

2.  Then we need to call the API

    ```{r}
    acsData <- "https://api.census.gov/data/2022/acs/acs1/pums?get=PWGTP,AGEP,SEX&for=region:04"
    ```

3.  Then we need to parse the JSON data and put it in tibble format with the helper function

    ```{r}
    apiHelper <- function(apiCall) {
      acsData <- GET(apiCall)
      parsed <- fromJSON(rawToChar(acsData$content))
      acsDataTib <- as_tibble(parsed)
      names(acsDataTib) <- acsDataTib[1,]
      acsDataTib <- acsDataTib[-1,]
    }
    ```

4.  Now were going to write a function to query the API that will allow the user to specify certain items. The function allows for the individual to select a year, what Numerical Variables they wish, Categorical Variables, and the Geography from the inputs into the function. How the function is written, causes only one singular "Geography" selection, while the Numerical and Categorical Variables are able to have multiple responses, as these are only the column headers being selected.

    ```{r}
    acsFunction <- function(year = 2022, 
                            numericalVar = c("PWGTP", "AGEP"), 
                            categoricalVar = "SEX", 
                            geography = "state",
                            geo_code = "04") {

      ## Checking if function input is valid
      if (year > 2022 | year < 2010) {
        stop("Year must be between 2010 and 2022")
      }
      
      if (is.null(categoricalVar)) {
        stop("One categorical variable must be returned")
      }
      
      if (is.null(numericalVar)) {
        stop("One numeric variable other than PWGTP must be returned")
      }
      
      if (!(geography %in% c("state", "division", "region"))) {
        stop("Invalid Geography value, must be: (state, division, region)")
      }
      
      if (is.null(geo_code)) {
        geo_code <- "*"
      }
      
      ## Building the URL
      variables <- paste(c(numericalVar, categoricalVar), collapse = ",")
      
      ## Ensures PWGTP is always printed
      if (!("PWGTP") %in% numericalVar) {
        variables <- paste("PWGTP", variables, sep = ",")
      }
      
      url <- paste0("https://api.census.gov/data/",
                    year,
                    "/acs/acs1/pums?get=",
                    variables,
                    "&for=",
                    geography,
                    ":",
                    geo_code
                    )
      
      acsTibble <- apiHelper(apiCall = url)
      
      ## Making variables numeric
      numericalVar <- numericalVar[!(numericalVar %in% c("JWAP", "JWDP"))]
      acsTibble[numericalVar] <- lapply(acsTibble[numericalVar], as.numeric)
      acsTibble$PWGTP <- as.numeric(acsTibble$PWGTP)
      
      ## Making variables factors
      acsTibble[categoricalVar] <- lapply(acsTibble[categoricalVar], as.factor)
      
      ## Formatting the time variables
      if ("JWAP" %in% numericalVar) {
        temp <- GET("https://api.census.gov/data/2022/acs/acs1/pums/variables/JWAP.json")
        temp_list <- temp$content |> rawToChar() |> fromJSON()
        temp_JWAP <- temp_list$values$item
        JWAP_values <- JWAP[sort(names(temp_JWAP))]
        JWAP_values[1] <- 0
        
        for (i in 2:length(JWAP_values)) {
          time_strings <- strsplit(JWAP_values[[i]], " to ")
          time_1 <- time_strings[[1]][1]
          time_2 <- time_strings[[1]][2]
          
          if (grepl("a\\.m\\.", time_1)) {
            time_form_1 <- gsub("a\\.m\\.", "AM", time_1)
          } else {
            time_form_1 <- gsub("p\\.m\\.", "PM", time_1)
          }
          
          if (grepl("a\\.m\\.", time_2)) {
            time_form_2 <- gsub("a\\.m\\.", "AM", time_2)
          } else {
            time_form_2 <- gsub("p\\.m\\.", "PM", time_2)
          }
          
          time_24_1 <- format(strptime(time_form_1, "%I:%M %p"), format="%H:%M")
          time_24_2 <- format(strptime(time_form_2, "%I:%M %p"), format="%H:%M")
          
          time_parts_1 <- strsplit(time_24_1, ":")[[1]]
          time_parts_2 <- strsplit(time_24_2, ":")[[1]]
          
          hours_1 <- as.numeric(time_parts_1[1])
          hours_2 <- as.numeric(time_parts_2[1])
          minutes_1 <- as.numeric(time_parts_1[2])
          minutes_2 <- as.numeric(time_parts_2[2])
          
          total_time_1 <- hours_1 * 60 + minutes_1
          total_time_2 <- hours_2 * 60 + minutes_2
          
          JWAP_values[i] <- (total_time_1 + total_time_2) / 2
        }
      }
      
      return(acsTibble)
    }
    ```

    ```{r}
    # Code intended for testing time variable conversion. will be deleted once project complete
    temp <- httr::GET("https://api.census.gov/data/2022/acs/acs1/pums/variables/JWAP.json")

    temp_list <- temp$content |> rawToChar() |> jsonlite::fromJSON()

    JWAP <- temp_list$values$item

    JWAP_values <- JWAP[sort(names(JWAP))]
    time_str <- strsplit(JWAP_values[[3]], " to ")
    time_test <- time_str[[1]][2]
    time_24 <- gsub("a\\.m\\.", "AM", time_test)

    formatted_time <- format(strptime(time_24, "%I:%M %p"), format="%H:%M")
    formatted_time

    time_parts <- strsplit(formatted_time, ":")[[1]]
    hours <- as.numeric(time_parts[1])
    minutes <- as.numeric(time_parts[2]) 
    hours * 60 + minutes
    average_time <- c()
    average_time[1] <- 0

    for (i in 2:length(JWAP_values)) {
      time_strings <- strsplit(JWAP_values[[i]], " to ")
      time_1 <- time_strings[[1]][1]
      time_2 <- time_strings[[1]][2]
      
      if (grepl("a\\.m\\.", time_1)) {
    time_form_1 <- gsub("a\\.m\\.", "AM", time_1)
      } else {
    time_form_1 <- gsub("p\\.m\\.", "PM", time_1)
      }
      
      if (grepl("a\\.m\\.", time_2)) {
    time_form_2 <- gsub("a\\.m\\.", "AM", time_2)
      } else {
    time_form_2 <- gsub("p\\.m\\.", "PM", time_2)
      }
      
      time_24_1 <- format(strptime(time_form_1, "%I:%M %p"), format="%H:%M")
      time_24_2 <- format(strptime(time_form_2, "%I:%M %p"), format="%H:%M")
      
      time_parts_1 <- strsplit(time_24_1, ":")[[1]]
      time_parts_2 <- strsplit(time_24_2, ":")[[1]]
      
      hours_1 <- as.numeric(time_parts_1[1])
      hours_2 <- as.numeric(time_parts_2[1])
      minutes_1 <- as.numeric(time_parts_1[2])
      minutes_2 <- as.numeric(time_parts_2[2])
      
      total_time_1 <- hours_1 * 60 + minutes_1
      total_time_2 <- hours_2 * 60 + minutes_2
      
      average_time[i] <- (total_time_1 + total_time_2) / 2
    }
    average_time
    ```

```{r}
    ## Test to show that the function works
    Result <- acsFunction(year = 2021, 
                        numericalVar = c("GASP", "GRPIP"),
                        categoricalVar = "HHL",
                        geography = "state",
                        geo_code = "06")
    
    Result2 <- acsFunction(geography = "division",
                           geo_code = "3")
```

```{r}
test <- acsFunction(numericalVar = c("JWAP","GASP"))
test
```

5.  Now we will build a new function that has similar parameters to the one in step 4 with the exception of allowing the user to specify multiple years. This will call the previous function multiple times and combine multiple tibbles into one.

```{r}
acsMultiYear <- function(years,
                         numericalVar = c("PWGTP", "AGEP"), 
                         categoricalVar = "SEX", 
                         geography = "state",
                         geo_code = "04") {
  tibbleList <- list()
  for (i in 1:length(years)) {
    singleYrTib <- acsFunction(years[i], numericalVar, categoricalVar, geography, geo_code)
    singleYrTib$year <- years[i]
    tibbleList[[i]] <- singleYrTib
  }
  multYrTib <- bind_rows(tibbleList)
  return(multYrTib)
}
```

```{r}
multYrTest <- acsMultiYear(years = c(2021, 2022))
head(multYrTest)
tail(multYrTest)
```

6.  In this step, we are creating a summary function of our overall tibble, while also adding a new class to the overall data. The summary is dependent on the numerical and categorical variables selected in the acsFunction previously.

    ```{r}

    summary.census<-function(
        classcensus = acsTibble,
        numericalVar = c("AGEP"),
        categoricalVar = "SEX"){
        
        #adding the class to the tibble#
        class(classcensus)<- c("census", class(classcensus))
        
        if (is.null(categoricalVar)) {
        stop("Atleast one categorical variable, that was selected in the acsFunction, must be returned")
      }
      
      if (is.null(numericalVar)) {
        stop("One numeric variable, that was selected in the acsFunction, other than PWGTP must be returned")
      }
        
        #Changing PWGTP to a factor to not be summarized#
        classcensus$PWGTP <- as.factor(classcensus$PWGTP)
        
        #Changing categorical variables back into numeric variables to be summarized#
        classcensus[categoricalVar] <- lapply(classcensus[categoricalVar], as.numeric)
        
        #Summary of the Tibble by PWGTP and Geographical selections with the mean and standard deviations#
        acsTibblesummary<-classcensus %>% 
          group_by(PWGTP) %>% 
          mutate(across(is.numeric, mean, .names = "mean_{.col}"),
                 across(is.numeric, sd, .names = "sd_{.col}")) %>% 
          select(PWGTP, any_of(c("state","region", "division")), starts_with("mean"),starts_with("sd_"),-starts_with("sd_mean_")) %>% 
          distinct()
        
        
        acsTibblesummary
      }
      
    ```

```{r}
 #test of the summary data#
testsumtibble<-acsFunction(year = 2021, 
                        numericalVar = c("GASP", "GRPIP"),
                        categoricalVar = "HHL",
                        geography = "state",
                        geo_code = "06")
testsummary<-summary.census(testsumtibble, 
                            numericalVar = c("GASP", "GRPIP"),
                            categoricalVar = "HHL")



```

7.  After creating our summary of our selected data, we are now able to plot the data in a digestible graph!

    ```{r}

    summaryplotfunct<-function(dataset = testsummary,
                      numericalVar = "mean_GASP",
                      categoricalVar = "mean_HHL"){
      
      dataset$PWGTP <- as.numeric(dataset$PWGTP)
      
      ggplot(dataset,
             aes({{categoricalVar}},
                 {{numericalVar}},
                 weight = PWGTP,
                 group = PWGTP)) +
        geom_boxplot()
      
    }
    ```

    ```{r}

    summaryplotfunct(testsummary, mean_GASP, mean_HHL)
    ```
